A continuous urbanization index to assess phylogenetic adaptations in urban bird living

### 2018-05-15 ms.Rmd
some minor odds and ends (keywords, formatting, refs, etc.)

author: |
  | Corey T. Callaghan^[Corresponding author: c.callaghan@unsw.edu.au] $^1$, Richard E. Major $^1$$^,$$^2$, John H. Wilshire $^1$, John M. Martin $^1$$^,$$^3$, Richard T. Kingsford $^1$, William K. Cornwell $^1$$^,$$^4$
  | $^1$Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia; $^2$Australian Museum Research Institute, Sydney, NSW, Australia; $^3$Royal Botanic Gardens and Domain Trust, Sydney, NSW, Australia; $^4$Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia
date: ""
output:
  rmarkdown::pdf_document:
    fig_caption: yes    
    includes:  
      in_header: preamble-latex.tex
fontsize: 12pt
link-citations: no
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ecology-letters.csl
bibliography: references.bib

---

@@

# INTRODUCTION
Urbanization fragments and isolates natural habitat [@marzluff2001restoration; @melles2003urban; @loram2007urban], drastically altering species' composition and ecosystem functions. Thus, urbanization globally threatens biodiversity through processes such as biotic homogenization [@mckinney2006urbanization; @mckinney2008effects] and is a major threat to conservation [@czech2000economic; @dearborn2010motivations]. Despite the documented loss of biodiversity directly or indirectly caused by urbanization [@mckinney1999biotic; @mckinney2006urbanization; @devictor2007functional], some species have successfully adapted to, colonized, and subsequently thrived in urban environments [@mckinney2002urbanization; @chace2006urban; @lowry2013behavioural; @mcdonnell2015adaptation]. For instance, some species which rarely occur within urban areas can be relatively common in surrounding rural habitats [@tratalos2007bird; @evans2011makes; @gatesire2014bird].

@@

## Phylogenetic analysis
For the phylogenetic portion of our analyses, we used the phylogenetic tree provided by Jetz et al. [@jetz2012global]. This phylogeny represents the first completely mapped phylogeny of 9,993 birds [@jetz2012global]. In order to assess whether there was a phylogenetic signal [@revell2008phylogenetic, @munkemuller2012measure] of the urbanization index and the continuous predictor variables used in our study, we used the phylosignal package [@keck2016phylosignal] in the R computing environment [@r2017coreteam]. This was also conducted for each of the continuous response variables in order to assess the relative contribution of phylogeny on the predictor variables.

@@

### Non-phylogenetically constrained models
Given that we hypothesized any of the potential candidate traits could significantly drive the response in our urbanization index, we first ran one global model including all 22 predictor variables. We then used the standardize function from the arm package [@gelman2008scaling; @gelman2016arm] to standardize model estimates to represent relative effect sizes. In order to confirm our linear model results, we employed a model-averaging approach [@grueber2011multimodel] in which we 'dredged' possible subsets of the global model. This was done using the dredge function in the MuMIn package [@barton2016mumin], and we allowed a total of 11 candidate predictors to be fit at any one time in order to avoid over-fitting. This left a total of 784,626 models which were assessed, and the model-averaged parameter estimates were calculated from the top models of those with ???AIC < 4. As with the global model, we present the standardized estimates which represent relative effect sizes

@@

## Urbanization index
For the 477 species included in the models the mean ? SD urbanization index was 0.91 ? 1.76, with a range from 0.008 (Black-backed Butcherbird) to 11.85 (Powerful Owl; Fig. 2).

@@

### Differences in non-phylogenetic and phylogenetic models
The phylogenetic and non-phylogenetic models largely agreed with the relative rank of standardized predictor estimates, and the significance of individual predictors (Fig. 4). However, brain residual size was not significant in either of the non-phylogenetic models, but was significantly associated with urbanization in our phylogenetic models (Fig. 5). Further, diet generalism, feeding habitat generalism, and breeding habitat generalism were relatively more important (larger parameter estimates) in the phylogenetic model compared with the non-phylogenetic model (Fig. 5). Lastly, although not significant, body size was minimally important for the non-phylogenetic model, but substantially more important for the phylogenetically controlled models (Fig. 4 & Fig. 5).

# DISCUSSION
Niche breadth, as measured by different types of generalism, was significantly associated with our urbanization index across both phylogenetic and non-phylogenetic analyses, confirming the notion that urban birds have broad environmental tolerance [@bonier2007urban]. Conversely, granivores and insectivores were consistently negatively associated with urbanization. Furthermore, we found a significant phylogenetic signal [@revell2008phylogenetic] of the urbanization index in our analysis, suggesting that there are phylogenetic constraints which allow certain birds to persist within urban environments [@morelli2016evidence, @ibanez2017global, @sol2017urbanisation, @la2018phylogenetic]. This contrasts with previous studies [@evans2011makes], but this contrasting finding is likely a result of increased statistical power from our increased sample size: 477 species in this study compared with 55 species in Evans *et al.* [-@evans2011makes]. Our results demonstrate that considering phylogeny in regards to this question is important.

The effect of treating species as non-independent by considering phylogeny in the context of the linear model was apparent. For example, there was an increase in parameter estimates of the most important parameters (e.g., clutch size and feeding habitat generalism; Figures 4 and 5), and this affect was robust relative to a consideration of phylogenetic uncertainty (Fig. S2). Further, residual brain mass (and body size to a lesser extent) was significantly important within the phylogenetic models, but not in the non-phylogenetic models, suggesting that within a clade the species with the largest residual brain mass and largest body size are most suited to surviving in urban areas [@major2010museum]. Conversely, other predictor variables became less crucial in the phylogenetically controlled models compared with the non-phylogenetic models, including some feeding types (i.e., insect specialists) and habitat preferences (i.e., agricultural). For those variables which became less important, it is likely that there may be particular clades that include a number of species that are associated with both that particular trait and urbanization [see @felsenstein1985confidence].

Our finding that ecological generalists are more readily observed in urban areas confirms that of other work [@mckinney2006urbanization, @bonier2007urban, @moller2009successful]. We also confirm work which has found a relationship between annual fecundity and urbanization in birds [@croci2008does, @moller2009successful]. While some studies have found a lack of evidence that relative brain size is critical in urban birds [@kark2007living, @evans2011makes], we confirm those which do support the hypothesis that brain size is heavily correlated with urban birds [@moller2009successful, @maklakov2011brains, @moller2015brain]. Furthermore, we do confirm that urbanization appears to select against insectivorous species [@chace2006urban, @evans2011makes]. While others have found that urbanization supports granivorous species [@chace2006urban, @kark2007living, @evans2011makes], we found the opposite trend, but this could be a result of relatively little supplementary feeding in Australian cities [cf., @fuller2008garden]. We did not find any strong support for our hypotheses on different nesting types, but we did find that in general, species which tended to be more social were at an advantage to city living, confirming previous studies [@kark2007living].

@@

Global broad-scale empirical datasets are rapidly expanding our knowledge of urbanization impacts on birds [e.g., @aronson2014global, @aronson2016hierarchical, @gutierrez2018citizen], in large part due to the growing use of citizen science data [@silvertown2009new] in combination with advances in remote sensing technology [@gorelick2017google]. We used a novel approach, over a broad spatial and temporal scale, to quantify an urbanization index in birds. We did not explicitly account for changes throughout the annual cycle in a species' association with urban areas [e.g., @la2014role, @lepczyk2017global, @katuwaleffect], but migratory status was considered as a predictor. We highlight that future studies, especially within migrant-dominated landscapes (e.g., Europe, North America), should account for changes in a species' response to urbanization throughout the full-annual cycle [@hostetler2015full, @marra2015call]. Further, future studies should look to expand this analysis outside the scope of Australia, and also investigate whether these patterns hold true at multiple spatial scales [@suarez2018scale] (i.e., does city size affect the patterns found here; [@sol2017urbanisation, @batary2018non]).

@@

# REFERENCES

@@

```{r Figure 4, echo=FALSE, fig.cap="Parameter estimates for each of the four modelling approaches employed in the analysis. A star represents a significant effect. Note that the model-averaged results do not contain every possible predictor variable in the case that a given predictor variable was not included in the top model set."}
knitr::include_graphics("finalFigs/Figure4.pdf")
```

>>>>

author: |
  | Corey T. Callaghan^[Corresponding author: c.callaghan@unsw.edu.au] $^1$, Richard E. Major $^1$$^,$$^2$, John H. Wilshire $^1$, John M. Martin $^1$$^,$$^3$, Richard T. Kingsford $^1$, William K. Cornwell $^1$$^,$$^4$
  | $^1$Centre for Ecosystem Science, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia; $^2$Australian Museum Research Institute, Sydney, NSW, Australia; $^3$Royal Botanic Gardens and Domain Trust, Sydney, NSW, Australia; $^4$Evolution and Ecology Research Centre, School of Biological, Earth and Environmental Sciences, UNSW Sydney, Sydney, NSW, Australia
date: ""
output:
  rmarkdown::pdf_document:
    fig_caption: yes    
    includes:  
      in_header: preamble-latex.tex
fontsize: 12pt
link-citations: no
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/ecology-letters.csl
bibliography: references.bib
---

@@

*keywords*: birds; brain size; generalists; specialists; life history traits; ecological traits; urbanization index

\newpage
# INTRODUCTION
Urbanization fragments and isolates natural habitat [@marzluff2001restoration; @melles2003urban; @loram2007urban], drastically altering species' composition and ecosystem functions. Thus, urbanization globally threatens biodiversity through processes such as biotic homogenization [@mckinney2006urbanization; @mckinney2008effects] and is a major threat to conservation [@czech2000economic; @dearborn2010motivations]. Despite the documented loss of biodiversity directly or indirectly caused by urbanization [@mckinney1999biotic; @mckinney2006urbanization; @devictor2007functional], some species have successfully adapted to, colonized, and subsequently thrived in urban environments [@mckinney2002urbanization; @chace2006urban; @lowry2013behavioural; @mcdonnell2015adaptation]. For instance, some species which rarely occur within urban areas can be relatively common in surrounding rural habitats [@tratalos2007bird; @evans2011makes; @gatesire2014bird].

@@

## Phylogenetic analysis
For the phylogenetic portion of our analyses, we used the phylogenetic tree provided by Jetz et al. [@jetz2012global]. This phylogeny represents the first completely mapped phylogeny of 9,993 birds [@jetz2012global]. In order to assess whether there was a phylogenetic signal [@revell2008phylogenetic; @munkemuller2012measure] of the urbanization index and the continuous predictor variables used in our study, we used the phylosignal package [@keck2016phylosignal] in the R computing environment [@r2017coreteam]. This was also conducted for each of the continuous response variables in order to assess the relative contribution of phylogeny on the predictor variables.

@@

### Non-phylogenetically constrained models
Given that we hypothesized any of the potential candidate traits could significantly drive the response in our urbanization index, we first ran one global model including all 22 predictor variables. We then used the standardize function from the arm package [@gelman2008scaling; @gelman2016arm] to standardize model estimates to represent relative effect sizes. In order to confirm our linear model results, we employed a model-averaging approach [@grueber2011multimodel] in which we 'dredged' possible subsets of the global model. This was done using the dredge function in the MuMIn package [@barton2016mumin], and we allowed a total of 11 candidate predictors to be fit at any one time in order to avoid over-fitting. This left a total of 784,626 models which were assessed, and the model-averaged parameter estimates were calculated from the top models of those with $\Delta$AIC < 4. As with the global model, we present the standardized estimates which represent relative effect sizes.

@@

## Urbanization index
For the 477 species included in the models the mean $\pm$ SD urbanization index was 0.91 $\pm$ 1.76, with a range from 0.008 (Black-backed Butcherbird) to 11.85 (Powerful Owl; Fig. 2).

@@

### Differences in non-phylogenetic and phylogenetic models
The phylogenetic and non-phylogenetic models largely agreed with the relative rank of standardized predictor estimates, and the significance of individual predictors (Fig. 4). However, brain residual size was not significant in either of the non-phylogenetic models, but was significantly associated with urbanization in our phylogenetic models (Fig. 5). Further, diet generalism, feeding habitat generalism, and breeding habitat generalism were relatively more important (larger parameter estimates) in the phylogenetic model compared with the non-phylogenetic model (Fig. 5). Lastly, although not significant, body size was minimally important for the non-phylogenetic model, but substantially more important for the phylogenetically controlled models (Fig. 4 & Fig. 5). Full model results for each of the four analyses can be found in Appendix 5.

# DISCUSSION
Niche breadth, as measured by different types of generalism, was significantly associated with our urbanization index across both phylogenetic and non-phylogenetic analyses, confirming the notion that urban birds have broad environmental tolerance [@bonier2007urban]. Conversely, granivores and insectivores were consistently negatively associated with urbanization. Furthermore, we found a significant phylogenetic signal [@revell2008phylogenetic] of the urbanization index in our analysis, suggesting that there are phylogenetic constraints which allow certain birds to persist within urban environments [@morelli2016evidence; @ibanez2017global; @sol2017urbanisation; @la2018phylogenetic]. This contrasts with previous studies [@evans2011makes], but this contrasting finding is likely a result of increased statistical power from our increased sample size: 477 species in this study compared with 55 species in Evans *et al.* [-@evans2011makes]. Our results demonstrate that considering phylogeny in regards to the association between ecological and life history traits and urbanization in birds is important.

The effect of treating species as non-independent by considering phylogeny in the context of the linear model was apparent. For example, there was an increase in parameter estimates of the most important parameters (e.g., clutch size and feeding habitat generalism; Figures 4 and 5), and this affect was robust relative to a consideration of phylogenetic uncertainty (Fig. S2). Further, residual brain mass (and body size to a lesser extent) was significantly important within the phylogenetic models, but not in the non-phylogenetic models, suggesting that within a clade the species with the largest residual brain mass and largest body size are most suited to surviving in urban areas [@major2010museum]. Conversely, other predictor variables became less crucial in the phylogenetically controlled models compared with the non-phylogenetic models, including some feeding types (i.e., insect specialists) and habitat preferences (i.e., agricultural). For the variables which became less important, it is likely that there may be particular clades that include a number of species that are associated with both that particular trait and urbanization [see @felsenstein1985confidence].

Our finding that ecological generalists are more readily observed in urban areas confirms that of other work [@mckinney2006urbanization; @bonier2007urban; @moller2009successful]. We also confirm work which has found a relationship between annual fecundity and urbanization in birds [@croci2008does; @moller2009successful]. While some studies have found a lack of evidence that relative brain size is critical in urban birds [@kark2007living; @evans2011makes], we confirm those which do support the hypothesis that brain size is heavily correlated with urban birds [@moller2009successful; @maklakov2011brains; @moller2015brain]. Furthermore, we do confirm that urbanization appears to select against insectivorous species [@chace2006urban; @evans2011makes]. While others have found that urbanization supports granivorous species [@chace2006urban; @kark2007living; @evans2011makes], we found the opposite trend, but this could be a result of relatively little supplementary feeding in Australian cities [cf., @fuller2008garden]. We did not find any strong support for our hypotheses on different nesting types, but we did find that in general, species which tended to be more social were at an advantage to city living, confirming previous studies [@coleman1994neophobia; @jokimaki1998distribution; @kark2007living].

@@

Global broad-scale empirical datasets are rapidly expanding our knowledge of urbanization impacts on birds [e.g., @aronson2014global; @aronson2016hierarchical; @gutierrez2018citizen], in large part due to the growing use of citizen science data [@silvertown2009new] in combination with advances in remote sensing technology [@gorelick2017google]. We used a novel approach, over a broad spatial and temporal scale, to quantify an urbanization index in birds. We did not explicitly account for changes throughout the annual cycle in a species' association with urban areas [e.g., @la2014role; @lepczyk2017global; @katuwaleffect], but migratory status was considered as a predictor. We highlight that future studies, especially within migrant-dominated landscapes (e.g., Europe, North America), should account for changes in a species' response to urbanization throughout the full-annual cycle [@hostetler2015full; @marra2015call]. Further, future studies should look to expand this analysis outside the scope of Australia, and also investigate whether these patterns hold true at multiple spatial scales [@suarez2018scale] (i.e., does city size affect the patterns found here; [@sol2017urbanisation; @batary2018non]).

@@

\newpage
# REFERENCES

@@

```{r Figure 4, echo=FALSE, fig.cap="Parameter estimates for each of the four modelling approaches employed in the analysis. A star represents a significant effect. Note that the model-averaged results do not contain every possible predictor variable in the case that a given predictor variable was not included in the top model set. Full model results can be found in Appendix 5."}
knitr::include_graphics("finalFigs/Figure4.pdf")
```
